---
title: "HW5_answer"
author: "Guojing Wu"
date: "3/7/2019"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = F, 
                      warning = F)
library(tidyverse)
library(pscl)
theme_set(theme_bw())
```

## Problem 1

### a) Fit a poisson model (M1)

```{r}
data_crab <- read.csv("HW5-crab.txt", sep = "")
fit.m1 <- glm(Sa ~ W, data = data_crab, family = poisson(link = log))
res.m1 <- summary(fit.m1)
```

The null model deviance $D_{null}$ is `r round(res.m1$null.deviance, 3)` while the M1 model deviance $D_{M1}$ is `r round(res.m1$deviance, 3)`, $D_{null} - D_{M1} \sim \chi_{1}^{2}$. Hence we got the pvalue = `r round(1 - pchisq(res.m1$null.deviance - res.m1$deviance, df = 1), 3)`, we reject the null hypothesis and state that the M1 model doesn't fit the data well.

We interpret that: the log rate ratio of number of satellites is `r round(res.m1$coefficients[2,1], 3)` given one unit change in carapace width.

### b) Fit a model (M2) with W and Wt, compare it with M1

```{r}
fit.m2 <- glm(Sa ~ W + Wt, data = data_crab, family = poisson(link = log))
res.m2 <- summary(fit.m2)
```

The M2 model deviance $D_{M2}$ is `r round(res.m2$deviance, 3)`, $D_{M1} - D_{M2} \sim \chi_{1}^{2}$. Hence we got the pvalue = `r round(1 - pchisq(res.m1$deviance - res.m2$deviance, df = 1), 3)`, we failed to reject the null hypothesis and state that the M2 model fits the data better.

We interpret that: 

* the log rate ratio of number of satellites is `r round(res.m2$coefficients[2,1], 3)` given one unit change in carapace width.

* the log rate ratio of number of satellites is `r round(res.m2$coefficients[3,1], 3)` given one unit change in weight.

### c) check over dispersion in M2

```{r}
residual.m2 = residuals(fit.m2, type = 'pearson', data = data_crab)
G = sum(residual.m2 ^ 2) # calc dispersion param based on full model
phi = G / (173 - 3)
plot(qnorm((173 + 1:173 + 0.5) / (2 * 173 + 1.125)), sort(abs(residual.m2)), xlab = 'Expected Half-Normal Order Stats', ylab = 'Ordered Abs Pearson Residuals')
abline(a = 0, b = 1)
abline(a = 0, b = sqrt(phi), lty = 2)
text(1, 4, labels = paste("phi = ", round(phi, 3), sep = ""))
```

The half normal plot shows that M2 does exist over dispersion.

```{r}
res.m2.dis <- summary(fit.m2, dispersion = phi)
```

We interpret that: 

* the log rate ratio of number of satellites is `r round(res.m2.dis$coefficients[2,1], 3)` given one unit change in carapace width.

* the log rate ratio of number of satellites is `r round(res.m2.dis$coefficients[3,1], 3)` given one unit change in weight.

## Problem 2

### a) 

```{r}
data_para <- read.csv("HW5-parasite.txt", sep = "") %>% 
  mutate(Year = as.factor(Year), 
         Area = as.factor(Area))
fit.para <- glm(Intensity ~ Area + Year + Length, data = data_para, family = poisson(link = log))
res.para <- summary(fit.para)
```

We interpret that: 

* the log rate ratio of intensity is `r round(res.para$coefficients[2,1], 3)` given area change from 1 to 2.

* the log rate ratio of intensity is `r round(res.para$coefficients[3,1], 3)` given area change from 1 to 3.

* the log rate ratio of intensity is `r round(res.para$coefficients[4,1], 3)` given area change from 1 to 4.

* the log rate ratio of intensity is `r round(res.para$coefficients[5,1], 3)` given year change from 1999 to 2000.

* the log rate ratio of intensity is `r round(res.para$coefficients[6,1], 3)` given year change from 1999 to 2001.

* the log rate ratio of intensity is `r round(res.para$coefficients[7,1], 3)` for 1 unit change in length.

### b) test for goodness of fit in (a)

```{r}
pval = 1 - pchisq(res.para$deviance, 1191 - 7)
```

The value we got is `r round(pval, 3)`, so we reject the null hypothesis and state that this model doesn't fit the data well.

### c) use zero-inflated model

We assume that the area may be realted with whether susceptible and build the model like this:

```{r}
fit.zip <- zeroinfl(Intensity ~ Area + Year + Length | Area, data = data_para)
res.zip <- summary(fit.zip)
```

We interpret for count part:

* within the susceptible fish, the log rate ratio of intensity is `r round(res.zip$coefficients$count[2,1], 3)` given area change from 1 to 2.

* within the susceptible fish, the log rate ratio of intensity is `r round(res.zip$coefficients$count[3,1], 3)` given area change from 1 to 3.

* within the susceptible fish, the log rate ratio of intensity is `r round(res.zip$coefficients$count[4,1], 3)` given area change from 1 to 4.

* within the susceptible fish, the log rate ratio of intensity is `r round(res.zip$coefficients$count[5,1], 3)` given year change from 1999 to 2000.

* within the susceptible fish, the log rate ratio of intensity is `r round(res.zip$coefficients$count[6,1], 3)` given year change from 1999 to 2001.

* within the susceptible fish, the log rate ratio of intensity is `r round(res.zip$coefficients$count[7,1], 3)` for 1 unit change in length.

We interpret for binomal part:

* the log odds ratio of susceptible or not is `r round(res.zip$coefficients$zero[2,1], 3)` given area change from 1 to 2.

* the log odds ratio of susceptible or not is `r round(res.zip$coefficients$zero[3,1], 3)` given area change from 1 to 3.

* the log odds ratio of susceptible or not is `r round(res.zip$coefficients$zero[4,1], 3)` given area change from 1 to 4.